Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-11 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_c4fec}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\subsection{Computing the Observed and Predicted Prevalence of
Involvement
Patterns}\label{computing-the-observed-and-predicted-prevalence-of-involvement-patterns}

@@

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to patients). Of those, were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is .

When displaying this data prevalence, we often choose to draw a
\emph{Beta posterior} over the ``true'' prevalence, hinting at the fact
that our data merely represents a limited sample. The Beta posterior
follows from a uniform Beta distribution as prior and a Bernoulli
likelihood for the number of patients with the involvement of interest,
given the parameer for the ``true'' prevalence. The resulting
distribution has its maximum at the observed prevalence, but in addition
gives a visual intuition for the variance of of the observed quantity.
I.e., when we observe 3 out of 10 events, the Beta posterior is much
wider than if we observe 300 out of 1000 for he same prevalence. It also
allows us to check not inly if the model is accurate, but also whether
it reflects the uncertainty contained in the data.

@@

\caption{}\label{T_b241f}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

\caption{}\label{T_bb1c4}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

\begin{itemize}
\tightlist
\item[$\square$]
  write short theory section about why and how to compute the posteriors
  if some of the data (here midline extension) is unknown
\end{itemize}

>>>>

\caption{}\label{T_42bd2}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\subsection{Computing the Observed and Predicted Prevalence of
Involvement Patterns}\label{sec-prevalence}

@@

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to 379 patients). Of those, 28 were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is
7\%.

When displaying this data prevalence, we often choose to draw a
\emph{beta posterior} over the ``true'' prevalence, hinting at the fact
that our data merely represents a limited sample. The beta posterior
follows from a uniform beta distribution as prior and a binomial
likelihood for the number of patients with the involvement of interest,
given the parameer for the ``true'' prevalence. The resulting
distribution has its maximum at the observed prevalence, but in addition
gives a visual intuition for the variance of of the observed quantity.
I.e., when we observe 3 out of 10 events, the beta posterior is much
wider than if we observe 300 out of 1000 for he same prevalence. It also
allows us to check not only if the model is accurate, but also whether
it reflects the uncertainty contained in the data.

@@

\caption{}\label{T_c1c15}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

\caption{}\label{T_78884}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

\begin{longtable}[]{@{}llll@{}}

\caption{\label{tbl-data-unknowns}Combinations of midline extension
status and state of contralateral LNL II for early T-category patients.}

\toprule\noalign{}
Midline Extension & no & yes & unknown \\
contra LNL II & & & \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
healthy & 352 & 23 & 59 \\
involved & 27 & 6 & 6 \\
total & 379 & 29 & 65 \\

In section~\ref{sec-prevalence} we explained that we compare the
predicted prevalence of an involvement pattern under a scenario to a
beta posterior over the ``true'' prevalence given the observations in
the data. However, our data contains incomplete observations, especially
regarding the tumor's midline extension. This information is not present
for a large group of 85 patients, most of which treated at the Centre
Léon Bérard in France.

To account for this missing information and plot the correct posteriors
over the data prevalence, we need to marginalize over the possible
combinations of matching involvement patterns and selected patients
under a scenario. For example, we have only 29 early T-category patients
with midline extension. But for an additional 65 patients we do not know
whether the tumor crossed the mid-sagittal plane. If we now want to
compute the prevalence of contralateral LNL II invovlement for these
early T-category patients with midline extension, then we need to
account for the possibility that we are missing information.

In table~\ref{tbl-data-unknowns} we show the combinations of the midline
extension status and state of the contralateral LNL II. Using only
available information, we would get to 6 out of 29 patients for the
prevalence of contralateral LNL II invovlement among the early
T-category patients with midline extension. But the column with the
``unknown'' numbers suggests that the true prevalence could range from 6
out of 29 + 59 (7\%) up to 12 out of 29 + 6 (34\%). Assuming that every
one of the possible combinations has the same a priori probability
(uniform prior) and accouting for permutations via the binomial factor,
the posterior marginalized over the unknowns could be written as

\[
\begin{aligned}
p(x; \alpha=1, \beta=1) &= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell} (1-x)^{n+m - (k+\ell)} \cdot \frac{1}{\operatorname{B}(\alpha, \beta)} x^\alpha (1-x)^\beta \\
&= \sum_{m=0}^{65} \sum_{\ell=0}^{\min(m,6)} \binom{n+m}{k+\ell} x^{k+\ell+1} (1-x)^{n+m - (k+\ell) + 1}
\end{aligned}
\]

We have plotted this marginal in figure~\ref{fig-model-marginal-pdf}
(blue curve) and the terms that make up the above sum (light red
curves). Note that we weighted the plotting opacity with the number of
permutations each of the possible combinations has. The marginal is
compared against the ``naive'' beta posterior that only takes the data
into account for which midline extension is available (green curve).

\includegraphics[width=0.5\textwidth,height=\textheight]{figures/fig-model-marginal-pdf.png}

\caption{\label{fig-model-marginal-pdf}Posterior distributions over the
data prevalence for contralateral LNL II involvement under the scenario
of early T-category and midline extension. The green beta distribution
shows this prevalence for those patients for which midline extension is
known. While the blue PDF is a marginal distribution summed over all
possible combinations, which are shown in shades of red. The more
permutations a particular combination has, the more opaque it is
plotted.}
